A Flux Scale for Southern Hemisphere 21cm EoR Experiments

### 2013-04-23 ms.tex
Pasted in unused text from psa32 power spectrum paper

Numerous radio telescopes are now exploring the prospects for using
measurements of highly redshifted 21cm emission to inform our understanding of
cosmic reionization in the redshift range $6< z<12$, corresponding to radio
frequencies below 200 MHz (see reviews in
\citealt{Furlanetto:2006p2267,Morales:2010p8093,Pritchard:2012p9555}) .  These
include telescopes aiming to measure the global temperature change of 21cm
emission during the Epoch of Reionization (EoR), such as the Compact
Reionization Experiment (CoRE) and the Experiment to Detect the Global EoR
Signature (EDGES; \citealt{Bowman:2010p8546}), and interferometers aiming to
measure the power spectrum of 21-cm EoR emission, such as the Giant Metre-wave
Radio Telescope (GMRT;
\citealt{Paciga:2011p9470,Paciga:2013p9627})\footnote{\url{http://gmrt.ncra.tifr.res.in/}},
the LOw Frequency ARray (LOFAR;
\citealt{Yatawatta:2013p9699})\footnote{\url{http://www.lofar.org/}}, the
Murchison Widefield Array (MWA;  \citealt{Bowman:2012p9138};
\citealt{Tingay:2013p9022}))\footnote{\url{http://www.mwatelescope.org/}}, and
the Donald C. Backer Precision Array for Probing the Epoch of Reionization
(PAPER; \citealt{Parsons:2010p6757,Parsons2013b})\footnote{\url{http://eor.berkeley.edu/}}.
Both PAPER and the MWA are located in the southern hemisphere at radio quiet
reserves being prepared for the upcoming Square Kilometer Array.  At these low
frequencies the Southern sky is much less well known than the North, with the
most extensive surveying work being conducted  by the EoR experiments
themselves \citep{Jacobs:2011p8438,Williams:2012p8768}.

Neutral hydrogen is visible during reionization as $\sim$30mK excess surface
brightness over the CMB punctuated both spectrally and spatially by regions of
zero emission.  This patchy signal is predicted to be visible everywhere on the
% XXX ARP: don't like description of EoR signal as "surface brightness"
sky at size scales up to degrees, or about $0.1$cMpc$^{-1}$.   To increase the
surface brightness sensitivity without building costly dishes, MWA, PAPER and
LOFAR have chosen to use combinations of phased and/or correlated dipoles. By
design also cover much wider fields of view ($>10$\arcdeg) and bandwidths
($\sim 100$\% fractional) than traditional dish telescopes.   As the elements
% XXX what about GMRT?
can not be physically pointed, the apparent flux scale varies across the sky.
This direction dependent gain is currently uncertain to 20\% or higher
\citep{Jacobs:2013p9713}.  The ideal solution is to use a wide-grid of
well-known calibrators to map the primary beam and set the flux scale.  
% XXX don't go to "ideal solution" just yet
% XXX ARP: add importance of accurate flux vs. frequency (Parsons et al. 2012, etc.)

@@

@@

@@

Antennae delays and amplitudes were found by fitting a point source visibility
model to Centaurus A, Pictor A, and Fornax A.  Each source is imperfectly
modeled by a single point-source, but the solution differences are minimized by
averaging over the three independent solutions. These same calibration
solutions have also been successfully applied in the power spectrum analysis by
Pober (2013) and Centaurus A imaging by Stefan (2013).

@@

@@

@@

\begin{figure}
\includegraphics[width=0.9\textwidth]{plots/PicA_Oct2011_uv_coverage.png}
\caption{The effective uv coverage, showing the relative contributions of each baseline when beam-forming on Pictor over a 9.6 hour synthesis in the October observing session. \label{fig:uv_coverage}}
\end{figure}

@@

The gain of each channel is the product of an overall instrumental passband and
the spectral dependence of the primary beam in the direction of the source.
Sources sharing the same track through the primary beam will have the same
average spectral gain function resulting from frequency dependent primary beam
attenuation.   We matched each measured source spectrum time sample to the
nearest sample (in altitude/azimuth) of a calibrator source
(J2331-416)\footnote{This source was chosen for the amount of data found in the
literature and the apparent smoothness of the spectrum.}.  We then, average the
two matched data streams in time.  Since both tracks make the same slice
through the primary beam, they will have the same average effective gain.
Assuming a power law model for J2331-416 (33Jy $\alpha=-0.76$), we compute a
bandpass gain.  After calibrating the raw 203 channel spectra with this
bandpass, we average into 10MHz sub-bands between 120 and 180 MHz, computing
the standard deviation as an estimate of the error.

>>>>

Numerous radio telescopes are now exploring the prospects for using
measurements of highly redshifted 21cm emission to inform our understanding of
cosmic reionization in the redshift range $6< z<12$, corresponding to radio
frequencies below 200 MHz (see reviews in
\citealt{Furlanetto:2006p2267,Morales:2010p8093,Pritchard:2012p9555}) .  These
include telescopes aiming to measure the global temperature change of 21cm
emission during the Epoch of Reionization (EoR), such as the Compact
Reionization Experiment (CoRE) and the Experiment to Detect the Global EoR
Signature (EDGES; \citealt{Bowman:2010p8546}), and interferometers aiming to
measure the power spectrum of 21-cm EoR emission, such as the Giant Metre-wave
Radio Telescope (GMRT;
\citealt{Paciga:2011p9470,Paciga:2013p9627})\footnote{\url{http://gmrt.ncra.tifr.res.in/}},
the LOw Frequency ARray (LOFAR;
\citealt{Yatawatta:2013p9699})\footnote{\url{http://www.lofar.org/}}, the
Murchison Widefield Array (MWA;  \citealt{Bowman:2012p9138};
\citealt{Tingay:2013p9022}))\footnote{\url{http://www.mwatelescope.org/}}, and
the Donald C. Backer Precision Array for Probing the Epoch of Reionization
(PAPER; \citealt{Parsons:2010p6757,Parsons2013b})\footnote{\url{http://eor.berkeley.edu/}}.

As such, there has been a renewed interest in the spectral and spatial variation of
foreground emission in the 100--200 MHz frequency band that covers
the $z=6$--13 redshift range expected to encompass reionization
\citep{furlanetto_et_al2006}.
In particular, the spectral properties of extra-galactic point-sources are important
both because they are valuable calibration references
and because they are strong foreground emitters that must be removed from 21-cm EoR measurements.
With the sparse availability of measured foreground properties in the
100--200-MHz frequency band over large areas of the sky \citep{deolivieracosta_et_al2008}, continued
foreground characterization is a
vital step en route to any 21-cm EoR detection.
Both PAPER and the MWA are located in the southern hemisphere at radio quiet
reserves being prepared for the upcoming Square Kilometer Array.  At these low
frequencies the Southern sky is much less well known than the North, with the
most extensive surveying work being conducted  by the EoR experiments
themselves \citep{Jacobs:2011p8438,Williams:2012p8768}.

% XXX is the "eor science" part necessary?  this isn't a paper on eor
Neutral hydrogen is visible during reionization as $\sim$30mK excess surface
brightness over the CMB punctuated both spectrally and spatially by regions of
zero emission.  This patchy signal is predicted to be visible everywhere on the
% XXX ARP: don't like description of EoR signal as "surface brightness"
sky at size scales up to degrees, or about $0.1$cMpc$^{-1}$.   To increase the
surface brightness sensitivity without building costly dishes, MWA, PAPER and
LOFAR have chosen to use combinations of phased and/or correlated dipoles. By
design also cover much wider fields of view ($>10$\arcdeg) and bandwidths
($\sim 100$\% fractional) than traditional dish telescopes.   As the elements
% XXX what about GMRT?
can not be physically pointed, the apparent flux scale varies across the sky.
This direction dependent gain is currently uncertain to 20\% or higher
\citep{Jacobs:2013p9713}.  The ideal solution is to use a wide-grid of
well-known calibrators to map the primary beam and set the flux scale.  
% XXX don't go to "ideal solution" just yet
% XXX ARP: add importance of accurate flux vs. frequency (Parsons et al. 2012, etc.)

@@

% XXX incorporate
%Given these shortcomings in the literature, we felt it prudent to examine-Pictor A directly.  We use data observed on JD2455747 
%from a 64-antenna single-polarization deployment of the 
%PAPER array in a minimum redundancy configuration that is more suited for imaging.  The details
%of these observations and their calibration are given in \citet{pober_et_al2013}.  In order to ensure that 
%errors in
%the beam model did not skew our result, we chose a set of calibrator sources near in declination
%to Pictor A that would follow nearly identical tracks through the primary beam.

% XXX incorporate
%To measure the spectrum of each source, we apply per-antenna phase and gain cal
%brations established-via self-calibration and remove
%the known dependence of amplifier gain and cable attenuation on ambient temperature was corrected
%using measured temperature data and temperature coefficients that were characterized in
%laboratory measurements \citep{parashare_bradley2009} and confirmed in field-measurements \citep{pober_et_al2012}.  For each source, we phase to the source 
%osition
%and sum over 
%all baselines greater than 20 wavelengths in length.
%A fringe-rate filter is applied to the resulting
%beam to suppress beam sidelobes where sources introduce variations that deviate more than 
%$\pm$0.1 mHz from the fringe rate of the source in question.

% XXX incorporate
%To derive a source spectrum from each drift-scan source profile, we compute a weighted average using
%weights derived from a model of
%PAPER's primary beam response as an effective approximation of inverse-variance weighting
%\citep{pober_et_al2012}, and averaging over the time series for each source.  Each output source spectrum
%represents a best estimate given the limitations of the beam model and flux calibration that were established
%up to this point.  To further improve these estimates, we derive an additional frequency-dependent gain parameter,
%smoothed on 10-MHz scales,
%that brings the spectrum of J2331-416 into closest agreement with a power-law spectrum of
%$S_{150}=31.9\ {\rm Jy}$, $\alpha=-0.69$ that represents the best fit to measurements below 1 GHz.
%This additional gain factor was applied to all measured spectra, including that
%of Pictor A.  The expectation is that, since all sources at similar declinations follow similar tracks 
%through the PAPER beam, any errors present in the original measured spectra as a result of differences between
%the simulated and actual beam response should be absent in the ratio between source spectra, and therefore
%should be corrected by the applied gain factor.

% XXX incorporate
%Finally, because Pictor A is resolved at the 20\% level on the longest (300m) East-West PAPER baselines, we 
%use per-baseline estimates of resolution weights to down-weight and correct for resolution effects.  To estimate
%resolution weights, we use the 90cm map of Pictor A presented in \citet{perley_et_al1997}.  Averaged over all
%the baselines in the array, the inclusion of resolution effects changed the measured flux by less than 5\%.

@@

% XXX merge with below
%Removed RFI by flagging samples in each frequency channel versus time that exceed the mean amplitude by 4-sigma
%after performing a time derivative.  Also flagged samples in each integrated spectrum that exceeded the
%mean amplitude across the spectrum by 6-sigma after performing a frequency derivative.  For efficiency, flagging
%was derived using only the subset of baselines involving a fiducial antenna.
%Frequencies below 120 MHz and above 180 MHz were also flagged, owing to roll-off of analog response.
%After RFI flagging, frequency channel resolution was decreased by a factor of four by summing channels.

@@

% XXX time-dependent calibration is critical.  modify paragraph below
Quantization gain through the correlator was linearized \citep{parsons_et_al2008}.
Finally, measurements of ambient temperature versus time near the balun amplifier of a fiducial antenna and near
receiver amplifiers were used to divide out the predicted gain variation versus temperature that these amplifiers
are known to exhibit.  A predicted gain profile was drived using coefficients characterized in
laboratory measurements \citep{parashare_bradley2009} and
confirmed in field measurements \citep{pober_et_al2011} of $H_{\rm balun}=-0.024$ dB/K, $H_{\rm recvr}=-0.045$
dB/K, and $H_{\rm cable}=-0.018$ dB/K for 150m-long cables.  Because cable temperatures were not measured
during these observations, the cable temperature was assumed to be the same as the measured balun temperature.

Antennae delays and amplitudes were found by fitting a point source visibility
model to Centaurus A, Pictor A, and Fornax A.  Each source is imperfectly
modeled by a single point-source, but the solution differences are minimized by
averaging over the three independent solutions. These same calibration
solutions have been successfully applied in \citet{Pober:2013XXX} and \citet{Stefan:2013:XXX}.
for power-spectrum analysis of foregrounds and imaging of Centaurus A, respectively.

% XXX differentiate amp cal here from spectral cal that comes later.

@@

In order to ensure that errors in
the beam model did not skew our result, we chose a set of sources near in declination
to Pictor A that would follow nearly identical tracks through the primary beam.
A list of these 
sources is given in Table \ref{tab:src_spec}, with the spectra measure by PAPER
illustrated in Figure \ref{fig:src_spec}.  These sources were chosen on the basis of brightness and
proximity in declination to Pictor A.
% XXX merge this with paragraph below

@@

% XXX incorporate paragraph below
%As the sky drifts through the primary beam of PAPER antennas, widely
%separated points will be observed with substantially different gains and
%so will require different data weightings to optimize SNR.
%Drift-scan observations of sources were divided by the calibrated beam response in the
%direction of the source as a function of time.  Spectra were calculated by averaging over
%time, weighting data at each time step by the square of the 150-MHz beam response in the source direction.
%Since noise and systematics relating to source observations are expected to be approximately constant
%versus time, while source amplitude is proportional to beam response, this weighting
%is an effective approximation of inverse-variance weighting \citep{pober_et_al2011}.

% XXX exclusion of minimum uv spacing?

@@

% XXX incorporate paragraph below
%To measure the spectrum of each source, we apply per-antenna phase and gain cal
%via self-calibration and remove
%the known dependence of amplifier gain and cable attenuation on ambient tempera
%using measured temperature data and temperature coefficients that were characte
%laboratory measurements \citep{parashare_bradley2009} and confirmed in field
%measurements \citep{pober_et_al2012}.  For each source, we phase to the source 
%and sum over 
%all baselines greater than 20 wavelengths in length.
%A fringe-rate filter is applied to the resulting
%beam to suppress beam sidelobes where sources introduce variations that deviate
%$\pm$0.1 mHz from the fringe rate of the source in question.

\begin{figure}
\includegraphics[width=0.9\textwidth]{plots/PicA_Oct2011_uv_coverage.png}
\caption{The effective uv coverage, showing the relative contributions of each baseline when beam-forming on Pictor over a 9.6 hour synthesis in the October observing session. \label{fig:uv_coverage}}
% XXX unclear whether amplitudes include our model of Pic A, or are even actual data vs. weights
\end{figure}

@@

% XXX selection of calibrator sources

The gain of each channel is the product of an overall instrumental passband and
the spectral dependence of the primary beam in the direction of the source.
Sources sharing the same track through the primary beam will have the same
average spectral gain function resulting from frequency dependent primary beam
attenuation.   We matched each measured source spectrum time sample to the
nearest sample (in altitude/azimuth) of a calibrator source
(J2331-416)\footnote{This source was chosen for the amount of data found in the
literature and the apparent smoothness of the spectrum.}.  We then, average the
% XXX clarify why one source used here, and more sources later 
two matched data streams in time.  Since both tracks make the same slice
through the primary beam, they will have the same average effective gain.
Assuming a power law model for J2331-416 (33Jy $\alpha=-0.76$), we compute a
bandpass gain.  After calibrating the raw 203 channel spectra with this
bandpass, we average into 10MHz sub-bands between 120 and 180 MHz, computing
the standard deviation as an estimate of the error.

% XXX merge this with above
%To further improve these estimates, we derive an additional frequency-dependent gain parameter,
%smoothed on 10-MHz scales,
%that brings the spectrum of J2331-416 into closest agreement with a power-law spectrum of
%$S_{150}=31.9\ {\rm Jy}$, $\alpha=-0.69$ that represents the best fit to measurements below 1 GHz
%\citep{slee1995,kuhr_et_al1981,large_et_al1981,burgess_hunstead2006}.
%This additional gain factor was applied to all measured spectra, including that
%of Pictor A.  The expectation is that, since all sources at similar declinations follow similar tracks 
%through the PAPER beam, any errors present in the original measured spectra as a result of differences between
%the simulated and actual beam response should be absent in the ratio between source spectra, and therefore
%should be corrected by the applied gain factor
